---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  gganimate = list(
    nframes = 50
  )
)
```

# AlleleShift 

`AlleleShift` is an R package to predict and visualize population-level changes in allele frequencies. 

The following manuscript provides details on the methods used by the package: Kindt R. 2021 (in prep.) AlleleShift: An R package to predict and visualize population-level changes in allele frequencies in response to climate change.

# Installation

The package can be installed from github via:

```{r, eval=FALSE}
# install.packages('devtools')
# devtools::install_github('RoelandKindt/AlleleShift')
```

The package has also been uploaded for review to CRAN after passing all the tests available via the [win-builder](https://win-builder.r-project.org/upload.aspx) web site. Should the package become available, then th recommended method of installation would be: `install.packages('AlleleShift')`. 


# Packages needed

```{r, warning=FALSE, message=FALSE}
library(BiodiversityR) # also loads vegan
library(poppr) # also loads adegenet
library(ggplot2)
library(ggsci)
library(ggforce)
library(dplyr)
library(ggrepel)
library(patchwork)
library(GGally)
library(mgcv)
library(ggmap)
library(gggibbous)
library(gganimate)

library(AlleleShift)

```




# Predicting shifts in allele frequencies

The following script is also available from the documentation of the 'count.model' function.

Input data 'Poptri.baseline.env' and 'Poptri.future.env' document bioclimatic conditions for populations in baseline and future climates. 'Poptri.genind' is information on allele counts of individuals in the 'adegenet::genind' format (note that packages 'adegenet' and 'poppr' offer various methods of importing data from other applications). 

```{r, warning=FALSE, message=FALSE}
# 1. Reduce the number of explanatory variables

data(Poptri.baseline.env)
data(Poptri.future.env)

VIF.select <- VIF.subset(Poptri.baseline.env, 
                         keep=c("MAT", "CMI"), 
                         cor.plot=TRUE)  

VIF.select$VIF$vars.included

baseline.env <- Poptri.baseline.env[, VIF.select$VIF$vars.included] 
summary(baseline.env)

future.env <- Poptri.future.env[, VIF.select$VIF$vars.included] 


# 2. Create the genpop object

data(Poptri.genind)
Poptri.genpop <- adegenet::genind2genpop(Poptri.genind)

# Get to know the populations and the alleles
poppr::poppr(Poptri.genind)
adegenet::makefreq(Poptri.genpop)

# 3. Calibrate the models

# Note that the ordistep procedure is not needed
Poptri.count.model <- count.model(Poptri.genpop, 
                                  env.data=baseline.env, 
				  ordistep=TRUE)

Poptri.pred.baseline <- count.pred(Poptri.count.model, env.data=baseline.env)
head(Poptri.pred.baseline)

Poptri.freq.model <- freq.model(Poptri.pred.baseline)
Poptri.freq.baseline <- freq.pred(Poptri.freq.model,
                                  count.predicted=Poptri.pred.baseline)

head(Poptri.freq.baseline)

# 4. Check how well the models predict baseline allele frequencies

# Populations are split in those with R2 > 0.50 and those with R2 < 0.50
# Better populations
plotA1 <- freq.ggplot(Poptri.freq.baseline, 
            plot.best=TRUE,
            ylim=c(0.0, 0.8))
plotA1

# Populations with low R2
manual.colour.values1 <- ggsci::pal_npg()(5)

plotB1 <- freq.ggplot(Poptri.freq.baseline, 
            plot.best=FALSE, 
            manual.colour.values=manual.colour.values1,
            xlim=c(0, 0.5),
            ylim=c(0, 0.25))
plotB1

# Colouring by alleles

plotA2 <- freq.ggplot(Poptri.freq.baseline, 
            colour.Pop=FALSE,
            plot.best=TRUE,
            ylim=c(0.0, 0.8),
            manual.colour.values=manual.colour.values1)

plotA2

plotB2 <- freq.ggplot(Poptri.freq.baseline, 
            colour.Pop=FALSE,
            plot.best=FALSE, 
            manual.colour.values=manual.colour.values1,
            xlim=c(0, 0.5),
            ylim=c(0, 0.25))

plotB2


# Note that you can also compare data with:
poppr::poppr(Poptri.genind)
adegenet::makefreq(Poptri.genpop)


# 5. Predict future allele frequencies

Poptri.pred.future <- count.pred(Poptri.count.model, env.data=future.env)
head(Poptri.pred.future)

Poptri.freq.future <- freq.pred(Poptri.freq.model,
                                count.predicted=Poptri.pred.future)

# The key results are variables 'Allele.freq' representing the baseline allele frequencies
# and variables 'Freq.e2', the predicted frequency for the future/ past climate.
# Variable 'Freq.e1' is the predicted allele frequency in step 1

head(Poptri.freq.future)

```

# Visualizations

## AlleleShift::population.shift

```{r, warning=FALSE, message=FALSE, fig1, fig.height=20/2.54, fig.width=17.2/2.54}
data(Poptri.baseline.env)
data(Poptri.future.env)

VIF.select <- VIF.subset(Poptri.baseline.env, 
                         keep=c("MAT", "CMI"), 
                         cor.plot=FALSE)  

VIF.select$vars.included

baseline.env <- Poptri.baseline.env[, VIF.select$vars.included] 
future.env <- Poptri.future.env[, VIF.select$vars.included] 

plotA <- population.shift(baseline.env, 
                 future.env,
                 option="PCA")
plotA

```

```{r, warning=FALSE, message=FALSE, fig2, fig.height=20/2.54, fig.width=17.2/2.54}

plotB <- population.shift(baseline.env, 
                 future.env,
                 option="RDA")
plotB
```






## AlleleShift::shift.dot.ggplot

```{r, warning=FALSE, message=FALSE, fig3, fig.height=20/2.54, fig.width=17.2/2.54}
# The data can be obtained via the count.model and freq.model calibrations.
# These procedures are not repeated here.
data(Poptri.freq.future)

ggdot1 <- shift.dot.ggplot(Poptri.freq.future)
ggdot1
```

## AlleleShift::shift.pie.ggplot

```{r, warning=FALSE, message=FALSE, fig4, fig.height=20/2.54, fig.width=17.2/2.54}
# The data can be obtained via the count.model and freq.model calibrations.
# These procedures are not repeated here.
data(Poptri.freq.baseline)
data(Poptri.freq.future)

Poptri.baseline.pie <- pie.baker(Poptri.freq.baseline, r0=0.1,
                                 sort.index="Latitude.index")

Poptri.future.pie <- pie.baker(Poptri.freq.future, r0=0.1, 
                        freq.focus="Freq.e2",
                        sort.index="Latitude.index",
                        ypos=1)

ggpie1 <- shift.pie.ggplot(Poptri.baseline.pie, 
                           Poptri.future.pie)
ggpie1
```

## AlleleShift::shift.moon.ggplot

```{r, warning=FALSE, message=FALSE, fig5, fig.height=20/2.54, fig.width=17.2/2.54}
# The data can be obtained via the count.model and freq.model calibrations.
# These procedures are not repeated here.
data(Poptri.freq.baseline)
data(Poptri.freq.future)

Poptri.baseline.moon <- moon.waxer(Poptri.freq.baseline,
                                   sort.index="Latitude.index")

Poptri.future.moon <- moon.waxer(Poptri.freq.future, 
                                 sort.index="Latitude.index",
                                 freq.focus="Freq.e2",
                                ypos=1)

ggmoon1 <- shift.moon.ggplot(Poptri.baseline.moon, 
                             Poptri.future.moon)
ggmoon1
```

## AlleleShift::shift.waffle.ggplot

```{r, warning=FALSE, message=FALSE, fig6, fig.height=20/2.54, fig.width=17.2/2.54}
# The data can be obtained via the count.model and freq.model calibrations.
# These procedures are not repeated here.
data(Poptri.freq.baseline)
data(Poptri.freq.future)

Poptri.future.waffle <- waffle.baker(Poptri.freq.future,
                           sort.index="Latitude.index")

ggwaffle1 <- shift.waffle.ggplot(Poptri.future.waffle)
ggwaffle1
```

## AlleleShift::shift.surf.ggplot

```{r, warning=FALSE, message=FALSE, fig7, fig.height=20/2.54, fig.width=17.2/2.54}
# The data can be obtained via the count.model and freq.model calibrations.
# These procedures are not repeated here.
data(Poptri.freq.baseline)
data(Poptri.freq.future)

# Plots for the first allele
# Symbols and colours indicate future change (green, ^ = future increase)
# Symbol size reflects the frequency in the climate shown

# Baseline climate
plotA <- shift.surf.ggplot(Poptri.freq.future,
                  xcoord="Long", ycoord="Lat",
                  Allele.focus=unique(Poptri.freq.future$Allele)[1], 
                  freq.focus="Allele.freq")
plotA

```

```{r, warning=FALSE, message=FALSE, fig8, fig.height=20/2.54, fig.width=17.2/2.54}

# Future/past climate
plotB <- shift.surf.ggplot(Poptri.freq.future,
                  xcoord="Long", ycoord="Lat",
                  Allele.focus=unique(Poptri.freq.future$Allele)[1], 
                  freq.focus="Freq.e2")
plotB

```



```{r, warning=FALSE, message=FALSE, fig9, fig.height=20/2.54, fig.width=17.2/2.54}

# Plots for the fifth allele

# Baseline climate
plotC <- shift.surf.ggplot(Poptri.freq.future,
                  xcoord="Long", ycoord="Lat",
                  Allele.focus=unique(Poptri.freq.future$Allele)[5], 
                  freq.focus="Allele.freq")
plotC

```


```{r, warning=FALSE, message=FALSE, fig10, fig.height=20/2.54, fig.width=17.2/2.54}


plotD <- shift.surf.ggplot(Poptri.freq.future,
                  xcoord="Long", ycoord="Lat",
                  Allele.focus=unique(Poptri.freq.future$Allele)[5], 
                  freq.focus="Freq.e2")
plotD
```



## Animations

With graphs that are generated via **ggplot2**, it is straightforward to create animations. The following example is also avaiable from the documentation of 'AlleleShift::shift.dot.ggplot'

```{r}
# create an animation

library(ggplot2)
library(gganimate)
library(gifski)

# The data is an interpolation and extrapolation between the baseline and future climate.
# For actual application, interpolate between climate data from available sources

data(Poptri.1985to2085)

ggdot.all <- ggplot(data=Poptri.1985to2085, group=Decade) +
  scale_y_continuous(limits=c(-0.1, 1.1),
                     breaks=c(0.0, 0.25, 0.5, 0.75, 1.0)) +
  geom_errorbar(aes(x=Pop, ymin=LCL, ymax=UCL),
                colour="grey30", width=0.8, show.legend=FALSE) +
  geom_segment(aes(x=Pop, y=Allele.freq, xend=Pop, yend=Freq.e2, colour=increasing),
               size=1.2) +  
  geom_point(aes(x=Pop, y=Allele.freq),
             colour="black", size=10, alpha=0.7) +                    
  geom_point(aes(x=Pop, y=Freq.e2),
             colour="dodgerblue3", size=10, alpha=0.7) +  
  coord_flip() +
  xlab(element_blank()) +
  ylab("Allele frequencies") +
  theme(panel.grid.minor = element_blank()) +
  labs(colour="Future change in allele frequencies") +
  scale_colour_manual(values=c("firebrick3", "chartreuse4"), 
                      labels=c("decreasing", "increasing")) +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, size=10)) +
  theme(legend.position="top") +
  facet_grid( ~ Allele, scales="free")

ggdot.all

ggdot.anim <- ggdot.all + 
  transition_states(as.factor(Decade), transition_length = 10, state_length = 100) +
  labs(title = "Decade: {closest_state}s")

```

Show the animation

```{r}
animate(ggdot.anim, fps=5, width=1280, height=720)
```






```{r}
sessionInfo()
```





